---
title: "Extra tables"
---

## Appendix

All tables and figures.

```{r}
#| include: false
#| output: false
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr")
if(!require(DT)) install.packages("DT")
library(plotly)
library(knitr)
library(DT)

library(data.table)
library(dplyr)
library(dtplyr)
library(tidyverse)
library(lubridate)
vac <- read.csv("vac.csv")
death <- read.csv("death.csv")
covid_income <- fread("income.csv", select = c("county", "AMI"))
density <- read.csv("density.csv")
```

```{r}
#| include: false
death$date <- as.Date(death$date, "%Y-%m-%d")
vac$administered_date <- as.Date(vac$administered_date,"%Y-%m-%d")

death <-
  death %>% mutate(county = area)

vac <-
  vac %>% mutate(date = administered_date)

vac1 <- vac
vac1$county <-replace(vac$county, vac$county=="All CA Counties","California")
```

```{r}
#| include: false
covid_merge = merge(x = death,y = vac1, by= c("date","county"), all=TRUE)
covid_merge <- covid_merge %>% 
  filter(!is.na(county))

covid_county <- 
  covid_merge %>%
  filter (county != "All CA and Non-CA Counties" & 
            county != "All CA Counties" &
            county != "California" &
            county !="Out of state"& 
            county !="Outside California"& 
            county !="Unknown")

covid_county <- covid_county %>%
  mutate(cum_dose_pp = (cumulative_total_doses/population),
         percent_vacc = (cumulative_fully_vaccinated/population))

covid_CA <- 
  covid_merge %>%
  filter (county == "California")

covid_county <- data.table(covid_county)
covid_county1 <- covid_county[!is.na(percent_vacc)]
covid_county1 <-
  covid_county1 %>%
  filter(date==as.Date("2023-05-09"))

covid_county1= merge(x=covid_county1, y=covid_income, by=c("county"),all=TRUE)
covid_county1= merge(x=covid_county1, y=density, by=c("county"),all=TRUE)
covid_county1 <- covid_county1 %>% mutate(pop_density = (population/Area))
```

### Table 1.

Summary Statistics for Counties in California (2020-02-01 to 2023-05-09)

```{r}
#| echo: false
#| warning: false
covid_county %>%
  group_by(county) %>%
  filter(date >= as.Date("2020-02-01")) %>%
  summarise(Cases_mean = mean(cases,na.rm = TRUE),
            Cases_max = max(cases,na.rm = TRUE),
            Cases_sd = sd(cases,na.rm = TRUE),
            Deaths_mean = mean(deaths,na.rm = TRUE),
            Deaths_max = max(deaths,na.rm = TRUE),
            Deaths_sd = sd(deaths,na.rm = TRUE),
            Population = mean(population,na.rm = TRUE),
            Dose_mean = mean(total_doses,na.rm = TRUE),
            Doses_max = max(total_doses,na.rm = TRUE),
            Doses_sd = sd(total_doses,na.rm = TRUE)) %>% 
  knitr::kable(col.names = c("County", 
                             "Mean Cases", 
                             "Max Cases", 
                             "SD Cases",
                             "Mean Deaths", 
                             "Max Deaths", 
                             "SD Deaths",
                             "Population", 
                             "Mean Doses", 
                             "Max Doses", 
                             "SD Doses"), digits = 1, "pipe", 
               caption = "Table 1. Summary Statistics for Counties in California (2020-02-01 to 2023-05-09)")
```

### Table 2a.

Top 5 vaccinated counties in California (as of 2023-05-09)

```{r}
#| echo: false
#| warning: false
covid_top5 <- covid_county1[rank(-percent_vacc) <=5]
covid_top5 <- covid_top5[order(-percent_vacc)]

covid_top5[,list(county,population,cumulative_cases,cumulative_deaths,cumulative_total_doses,percent_vacc,AMI)]%>% 
  knitr::kable(col.names = c("County", 
                             "Population", 
                             "Cumulative Cases", 
                             "Cumulative Deaths",
                             "Cumulative Total Doses", 
                             "Percent vaccinated", 
                             "AMI"), digits = 1, "pipe", 
               caption = "Table 2a. Top 5 vaccinated counties in California (as of 2023-05-09)")
```

### Table 2b.

Top 5 least vaccinated counties in California (as of 2023-05-09)

```{r}
#| echo: false
#| warning: false
covid_bot5 <- covid_county1[rank(-percent_vacc) >=54]
covid_bot5 <- covid_bot5[order(percent_vacc)]
covid_bot5[,list(county,population,cumulative_cases,cumulative_deaths,cumulative_total_doses,percent_vacc,AMI)]%>% 
  knitr::kable(col.names = c("County", 
                             "Population", 
                             "Cumulative Cases", 
                             "Cumulative Deaths",
                             "Cumulative Total Doses", 
                             "Percent vaccinated", 
                             "AMI"), digits = 1, "pipe", 
               caption = "Table 2b. Top 5 least vaccinated counties in California (as of 2023-05-09)")
```

### Figure 1.

Percent Fully Vaccinated by 2023-05-09 (ranked by County)

```{r fig.height=7}
#| echo: false
#| warning: false
covid_county1 %>%
  ggplot(aes(y = reorder(county, percent_vacc), x = percent_vacc)) +
  geom_bar(stat = "identity") +
  labs(y = "County", x = "Percent Fully Vaccinated by 2023-05-09") +
  theme(legend.position = "none") +
  ggtitle("Figure 1. Percent Fully Vaccinated by 2023-05-09 (ranked by County)")
```

### Figure 2.

Cumulative Total Doses per Population

```{r fig.height=6}
#| warning: false
#| echo: false
library(ggplot2)
covid_county %>%
  ggplot(mapping=aes(x = date, y = cum_dose_pp, color = county, na.rm = TRUE)) +
  geom_line() + geom_text(data=subset(covid_county, date=="2023-05-09"), aes(label=county),hjust=-.1) +
  scale_x_date(name="Date", limits = as.Date(c("2020-12-09", "2023-09-09"))) +
  scale_y_continuous(name="Cumulative Total Dose per Population", breaks = seq(0,3,by=1)) + 
  theme(legend.position="none") +
  ggtitle("Figure 2. Cumulative Total Doses per Population")
```

### Figure 3.

Daily Cases, Deaths, Doses from 2020-01-01 to 2023-05-13

```{r}
#| warning: false
#| echo: false
ax <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE
)
fig1 <- plot_ly(covid_CA, x=~date, y=~cases, name="cases", type="scatter", mode="lines") %>%
  layout(xaxis = ax, yaxis = list(title = 'Cases'))

fig2 <- plot_ly(covid_CA, x=~date, y=~deaths, name='deaths',type="scatter", mode="lines")%>%
  layout(xaxis = ax, yaxis = list(title = 'Deaths'))

fig3 <- plot_ly(covid_CA, x=~date, y=~total_doses, name='total doses',type="scatter", mode="lines") %>% layout(xaxis = list(title = 'Date'), yaxis = list(title = 'Doses'))

fig <- subplot(fig1, fig2, fig3, nrows=3,heights=c(0.25,0.2,0.55),titleY = TRUE, titleX = TRUE)%>%
  layout(hovermode="x unified", 
         title="Figure 3. Daily Cases, Deaths, Doses from 2020-01-01 to 2023-05-13")
fig
```

### Figure 4.

Population Density vs. Cummulative Cases by 2023-05-09

```{r}
#| echo: false
#| warning: false
covid_county1 %>% 
  plot_ly(x = ~pop_density, y = ~cumulative_cases, 
          type = "scatter", mode = 'markers', color = ~county,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))%>%
  layout(title="Figure 4. Population Density vs. Cummulative Cases by 2023-05-09",
         xaxis = list(title = 'Population density'), 
         yaxis = list(title = 'Cummulative cases'))
```

### Figure 5.

Area Median Income by Percent Fully Vaccinated by 2023-05-09

```{r}
#| echo: false
#| warning: false
fv <- covid_county1 %>% lm(percent_vacc~AMI,.) %>% fitted.values()
  
plot_ly(covid_county1, x = ~AMI, y = ~percent_vacc, mode="markers", text=~county) %>%
  add_markers(y = ~percent_vacc) %>%
  add_trace(x = ~AMI, y = fv, mode="lines") %>%
  layout(title="Figure 5. Area Median Income by Percent Fully Vaccinated by 2023-05-09",
         xaxis = list(title = 'Area Median Income ($)'), 
         yaxis = list(title = 'Percent Fully Vaccinated'),
         showlegend=FALSE)
```
